---
title: "The Political Elite & The Welfare State"
author: "Bas Machielsen"
institute: "Utrecht University <br> <br>a.h.machielsen@uu.nl &nbsp; <a href='http://github.com/basm92'> github.com/basm92 </a>"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE, warning = FALSE}
library(xaringanthemer)

style_duo(
    primary_color = "#1F4257", 
    secondary_color = "#F97B64",
    header_font_google = google_font("Josefin Sans"),
    text_font_google   = google_font("Piazzolla"),
    text_font_size = "1.2em"
    )

```

```{r, load_refs, include=FALSE, cache=FALSE}
library(RefManageR)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
mybib <- ReadBib("./mybib.bib", check = FALSE)
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: center, middle, inverse

# Motivation

---

# Motivation

- Many (Western) European countries underwent radical changes between roughly 1850 and 1920. 
    - All of them experienced high economic growth and were rapidly industrializing. 
    - Most of them transitioned to universal suffrage
    - Some of them established the groundwork of social security.

- At the same time, the countries that took part in this trajectory started as being governed by an oligarchical elite. 

- Why did these countries choose to go down this trajectory, and what is the role in this development of politicians?

---

# Motivation

- Roughly speaking, two kinds of changes took place:

    - **Increasing government intervention**:
    
    In many countries, governments sowed the seeds of the welfare state. Public spending increased, education was being publicly funded, the first pensions came into existence, as did public insurances (Lindert, 2002). Of course, simultaneously, increases in the government budget and the tax burden occurred. 
    
    - **Suffrage extension and democratization**:
    
    Governments repeatedly decided to extend the suffrage (Przeworski, 2002). In many cases, democratization did not happen overnight, but gradually. Almost all Western European countries established universal suffrage between roughly 1900 and 1930. 
    
---

# Motivation

- In this paper, I want to concentrate on the role of **politicians**, and more specifically, the role of **self-interest** in the decision-making leading to the aforementioned trends (**democratization** and **increasing government intervention**)

- The role of politicians has been researched in various settings, and various aspects of politicians have been investigated.
    
    - For example, (political) historians tend to focus on the ideology and personal experiences of politicians. 
    - On the other hand, political scientists and economists tend to focus on the electoral incentives politicians are faced with. 
    
- Politicians' **self-interest** is the key factor that is suggested theoretically by both groups of researchers, but has been left out of focus empirically. 

---

# Motivation

- I investigate the role of **self-interest** of politicians in a country that is much like other Western European countries in many aspects in the $19^{\text{th}}$ century: **the Netherlands**. 

    - The Netherlands oversaw a trend from a very partially democratic society in the 1850's to universal suffrage in 1919. 
    - It saw an increase in government spending, and an increasing in tax burden, particularly in the 1910's. 
    - In this respect, it is much like (e.g.) Germany, Denmark, the United Kingdom, Sweden, or Austria. 

- In other aspects, the Netherlands is less like aforementioned countries.
    - The late 19th century saw an increase in religious tensions and lead to a situation called **pillarization**, which has been argued to be a Dutch particularity (Lijphart, 2008; De Rooy, 2001) 

---

class: center, middle, inverse

# Literature 

---

# Literature

- In the literature, different disciplines have tried to explain trends towards **democratization** and towards the **welfare state** or **increased government spending**. 

- The **Dutch politico-historical literature** stresses aspects such as:
    - **Ideology**: De Haan (2003) - Confessionals and liberals found themselves united in opposition against an expanding state. An ideological turn among liberals caused them to make a U-turn and attempt collaboration with socialists.
    - **Pacification**: Lijphardt (2008) stresses the role of political **leadership** and the reputation of leaders among their electoral base. 
    - **Balance of power**: Van Zanden and Van Riel (2004) argue that the **constitution** established the fundamentals for inclusive economic growth. 
    - But also: **personal interests**. Van den Berg (1983), Secker (1991) explicitly claim that personal interests (such as wealth) and other demographic characteristics influence politicians' decisions. 

---

# Literature

- The **political science & economics** literature:
    - Demographic and environmental factors:
        - Aidt and Dallal (2008): **Women's suffrage** has a positive influence on government spending.
        - Lacroix et al. (2019): **Political dynasties** influence the propensity to vote against subversion of democracy (France, 1940).
            
    - Electoral pressure: 
        - Aidt and Jensen (2009) on **adoption of income taxes**
        - Aidt and Franck (2019) on the role of **peaceful and violent unrest** on the 1832 reform bill in the UK. 
            
---

# Literature

- But also in this literature: personal interests. 
    - Mares and Queralt (2020): **Intra-elite competition** for power
    - Acemoglu et al. (2015): **Captured democracy** by political elites
    - Corvalan et al. (2018): Removal of **eligibility requirements**, but not suffrage requirements, leads to the election of less wealthy politicians and the enactment of more redistributive policies.

- Engerman and Sokoloff (2005) argue that the differences trajectories that lead to prosperity and democratization are mainly due to:
    > systematic differences in the ability and inclination of elites to shape legal frameworks to advantage themselves
        
---

# Gaps

- The empirical literature suffers from **availability bias**
    - The influence of **personal interests** is readily implied by theory, but very difficult to measure
    - There is evidence that personal wealth influences decision-making in a contemporary context (Tahoun and Van Lent, 2018)

- The extension of the franchise is considered to be **exogenous** in many studies, but this is **inconsistent**
    - Implied by theory on franchise extension (Lizzeri and Persico, 2004) and political selection (Besley, 2008) that this is not the case. 
    - No explanation as to why politicians would extend the franchise if they knew this would harm their own interests. 
    - Any analysis should tackle these interrelated questions (**franchise extensions, taxes and government intervention**) at once, and explain why and how the findings are consistent. 

---

class: center, middle, inverse

# Methodology & Data

---

# Methodology

- My principal research question is about the influence of politicians' self-interest on their voting behavior in **various important votes** that lead to the **acceptance** or **rejection** of various far-reaching laws throughout the period **1870-1920**. 
    - Votes in both the upper house (_Eerste Kamer_) and the lower house (_Tweede Kamer_)

- More precisely, I attempt to identify the influence of politicians' **personal wealth** on their voting behavior. 

- By paying close attention to the theory, I attempt to incorporate **as many insights** in the behavior of politicians **as possible**. 
    - I try to measure both factors suggested by political historians, as well as factors suggested by economists and political scientists. 


---

name: methodologyslide

# Methodology

- What do I do specifically?

    - I collect data on votes on a number of important laws and law projects [[here]](#listoflaws) identified as such by expert historians. 
        - Laws instigating **increased government intervention**
        - Its **financing by taxes**
        - And **suffrage extension**
    - I also collect data on the **wealth at death** of politicians from the _Memories van Successie_. 
    - I attempt to estimate in *various ways* the wealth of politicians at the **time of voting** on a **particular law**. 
    - From a wide array of sources, I collect data on various **control variables** to measure factors suggested by the **literature**. 


---

# Methodology

- The baseline empirical models that I try to estimate is, for each category $j \in \{\text{Taxes, Government Intervention, Suffrage Extension\}}$:

$$V_{i,k} = f(\alpha + \beta_1 \cdot \text{Wealth}_{i,k} + \beta_2 \cdot \text{LawDum}_{k} \cdot \text{Wealth}_{i,k} + \beta_3 \cdot \text{Controls}_{i,k} + \epsilon_{i,k})$$
- where the subscript $i$ denotes observations by a particular politician on a law $k$ within category $j$, and $f$ is the logistic function. 

    - In other words, I **pool** laws per category and perform logit regressions.

- I investigate both the unconditional influence of the wealth on voting in favor of **taxation**, **suffrage extension** and **government intervention**, as well as the unique effect per-law by the coefficients under $\beta_2$. 

- Main identification problem: endogeneity of **wealth at death**

---

# Methodology

- What do I do to __mitigate__ this problem?

- Calculate wealth at the time of voting from wealth at death:

    - $\text{Wealth}_{i,t+1}= \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j, [t, t+1]}$
    - Since I know **wealth at death**, and the **portfolio composition**, calculate the (average) real returns on asset class $j$ and estimate the wealth at the time of voting.
    - **Historical returns** provided by Jorda et al. (2019)
    
- Condition the analysis on politicians that died relatively **recently after voting**

    - Mitigates the endogeneity of wealth and voting behavior (Fisman et al., 2004)
    
- Instrumental variable estimation

---


# Data

- I combine various sources:

- Voting results: [Staten-Generaal Digitaal](http://www.statengeneraaldigitaal.nl) is the official database of the Dutch Estates-General, on which all minutes (_Handelingen_) of lower house and upper house sessions are available.
    - From these, I find the voting results of all laws under investigation. 
    
- Personal wealth: _Memories van Successie_: **probate inventories** administered by the Dutch agency, contains an appraisal of an individual's assets and liabilities with the purpose of levying **inheritance taxes**. 
    - Wealth returns: from the [Jorda-Schularick-Taylor Database](http://www.macrohistory.net/data/)

---

# Data

- Controls:

    - Data regarding politicians' careers, district representation, ideological and party affiliation, and social origin from the [Politiek Documentatie Centrum](http://www.pdc.nl)
    - Data on electoral controls: turnout, nearest competitor, newspaper recommendation, etc.  (for lower house members) from [this repository by Ron De Jong](http://resources.huygens.knaw.nl/verkiezingentweedekamer)
    - Data on the frequency and location of strikes from the [IISG Labour Conflicts database](https://collab.iisg.nl/web/labourconflicts/datafiles)
    - Data on demographic and economic characteristics of districts from a [database administered by Onno Boonstra](https://datasets.iisg.amsterdam/dataverse/HDNG?q=&types=files&sort=dateSort&order=desc&page=1) <sup>1</sup>
        - This database contains information on the municipality-level, which I aggregate to the district level (if applicable)
    
.footnote[
[1] I created an R package to combine the data and make it useable [here](https://github.com/basm92/hdng/)
]
    
---

class: center, middle, inverse 

# Analysis

---

# Analysis

---

# Analysis

---

# Analysis

---

class: center, middle, inverse

# Conclusion

---

# Conclusion

- Interpretation: on the whole, a nuanced, but **coherent** picture emerges.

- I found that politicians did **not** let their **personal interests** speak when deciding to extend the franchise and abolish eligibility requirements. 
   - In the rejected 1892 proposal to extend suffrage, there is evidence that richer politicians were more likely to vote *against* the suffrage extension.
   - However, the effect of **political ideology** was 10 times larger in magnitude. 
   - Apart from ideology, it is hard to see what caused them to extend the suffrage.
   - The findings **contradict** some findings from other countries (e.g. Aidt and Franck, 2019), which imply electoral incentives play a large role in reforming political institutions. 
   - Findings could be **specific to the Netherlands** (i.e. _pillarization_ could have played a large role beyond demographic factors).
   - Further research is necessary to find out what were the primary drivers of suffrage extension are.
   
---

# Conclusion

- I found that politicians' **personal interests** are **moderately negatively correlated** with the tendency to vote in favor of **increased government intervention**. 

    - For some laws, the effect is very strong, and the point estimate **highly significant**, but in others, it is not.
    - 
    - **Almost always**, the point estimate is in the **hypothesized direction**. 
    - Qualitative research can shed light on the precise contents of each laws. 
    - Possibly, amendments to the laws might radically change their spirit, and hence influence the decision to vote against the final project. 

---

class: center, middle, inverse

# Thanks for your attention!


Slides created via the R package [**<span style="color:#1F4257">xaringan</span>**](https://github.com/yihui/xaringan).


---

# References

```{r refs, echo=FALSE, results="asis", warning = FALSE}
PrintBibliography(mybib, start = 1, end = 7)
```

---

name: listoflaws

# List of Laws 

- Suffrage Extensions

```{r listoflaws, echo=FALSE}
library(knitr); library(kableExtra)

listoflaws_suffrage <- data.frame(Name = c(
        "Reforms of 1872",
        "Reforms of 1887",
        "Reforms of 1892",
        "Reforms of 1896",
        "Universal male suffrage", 
        "Universal suffrage"),
        Year = c("1872", "1887", "1892", "1896", "1917", "1919"),
        Status = c("Rejected", "Accepted", "Rejected", "Accepted", "Accepted", "Accepted"))

knitr::kable(listoflaws_suffrage) %>%
    column_spec(column = c(1,2,3), color = "black") %>%
    row_spec(row = 0, color = "black") %>%
    kable_styling(full_width = FALSE, bootstrap_options = "hover")
```


.footnote[
[1] [Go back](#methodologyslide)
]

---

# remark.js vs xaringan

Some differences between using remark.js (left) and using **xaringan** (right):

.pull-left[
1. Start with a boilerplate HTML file;

1. Plain Markdown;

1. Write JavaScript to autoplay slides;

1. Manually configure MathJax;

1. Highlight code with `*`;

1. Edit Markdown source and refresh browser to see updated slides;
]

.pull-right[
1. Start with an R Markdown document;

1. R Markdown (can embed R/other code chunks);

1. Provide an option `autoplay`;

1. MathJax just works;<sup>*</sup>

1. Highlight code with `{{}}`;

1. The RStudio addin "Infinite Moon Reader" automatically refreshes slides on changes;
]

.footnote[[*] Not really. See next page.]


---
